---
title: Getting Started with genpca
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting Started with genpca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = TRUE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```




## Fit a weighted analysis

`genpca()` takes a numeric matrix with **observations in rows and variables
in columns**. A row metric `M` weights observations; a column metric `A`
weights variables. With identity metrics and the same preprocessing, it
recovers ordinary PCA.

Start with the three crime variables in `USArrests`: 50 states by 3 variables.
Their spreads differ substantially. Set `A` to inverse column variances so
that Assault does not dominate simply because of its numerical scale.
This is equivalent to PCA on standardized columns.

```{r quick-start}
data("USArrests")
X <- as.matrix(USArrests[, c("Murder", "Assault", "Rape")])
col_sd <- apply(X, 2, sd)
A <- Diagonal(x = 1 / col_sd^2)
fit <- genpca(X, A = A, ncomp = 2,
              preproc = multivarious::center())

S <- multivarious::scores(fit)       # 50 states x 2 components
V <- multivarious::components(fit)   # 3 variables x 2 projection weights
round(cor(X, S), 2)
```

The variable-score correlations give the axes an interpretation. PC1 moves
with all three crime variables; PC2 contrasts Murder with Rape more strongly
than it contrasts Assault with Rape. Component signs are arbitrary: reversing
an axis leaves the analysis unchanged.

```{r quick-plot, echo = FALSE, fig.cap = "States in the two-component space after inverse-variance weighting of the crime variables. Six states with extreme scores are labelled.", fig.height = 4.5}
plot(S, pch = 19, col = "grey50",
     xlab = "PC1", ylab = "PC2", main = "Standardized crime variables",
     xlim = extendrange(S[, 1], f = 0.15),
     ylim = extendrange(S[, 2], f = 0.15))
label <- unique(c(order(S[, 1])[c(1, 2, 49, 50)], which.min(S[, 2]), which.max(S[, 2])))
text(S[label, 1], S[label, 2], labels = state.abb[label], pos = 3, cex = 0.8)
```

To see the effect of scaling, compare each variable's correlation with PC1
under the two fits. Absolute correlations remove the arbitrary axis sign.

```{r scaling-comparison}
fit_raw <- genpca(X, ncomp = 2, preproc = multivarious::center())
round(cbind(
  unscaled = abs(cor(X, multivarious::scores(fit_raw)[, 1])),
  scaled   = abs(cor(X, S[, 1]))
), 2)
```

```{r quick-check, include = FALSE}
stopifnot(identical(dim(S), c(50L, 2L)), all(is.finite(S)),
          all(abs(cor(X, S[, 1])) > 0.7))
```

### Use the fitted object

The result is a `bi_projector`. Use its public accessors to inspect the fit,
reconstruct the input, or project new rows with the training preprocessing.

```{r object-flow}
Xhat <- multivarious::reconstruct(fit)   # 50 x 3, back in original units
scores_again <- multivarious::project(fit, X)
max(abs(S - scores_again))              # numerical zero on training rows
```

| Operation | Output |
|:--|:--|
| `multivarious::scores(fit)` | `n x k` training coordinates |
| `multivarious::components(fit)` | `p x k` projection weights |
| `multivarious::project(fit, X_new)` | `n_new x k` coordinates, with training centering applied |
| `multivarious::reconstruct(fit, comp = 1:k)` | `n x p` reconstruction, with centering reversed |

New data must have the same variables in the same order. Centering is fitted
on the training data; do not center a new batch separately before `project()`.

## Add observation weights

A diagonal `M` changes how much each row contributes to the objective.
As an illustration, give greater weight to states with a higher urban
population percentage. This is an analytical choice, not a claim that those
states are more reliable or a weighting by total population.

```{r row-weights}
urban_wt <- USArrests$UrbanPop / mean(USArrests$UrbanPop)
M <- Diagonal(x = urban_wt)
fit_row <- genpca(X, M = M, A = A, ncomp = 2,
                  preproc = multivarious::center())
round(cor(X, multivarious::scores(fit_row)), 2)
```

The weights change the fitted axes; they do not push a state toward a
particular signed end of an axis. This fit retains ordinary column centering
through `multivarious::center()`; setting `M` does not change that
preprocessing into weighted centering.

## Choose the number of components

`fit$sdev` contains singular values of the metric-whitened data. With identity
metrics and centering, these equal `prcomp(X)$sdev * sqrt(nrow(X) - 1)`.
Fit the full available spectrum before using a scree plot to choose a rank:

```{r scree, fig.cap = "Fraction of total weighted variation for all three available components.", fig.height = 3}
fit_all <- genpca(X, A = A, ncomp = ncol(X),
                  preproc = multivarious::center())
share <- fit_all$sdev^2 / sum(fit_all$sdev^2)
barplot(share, names.arg = paste0("PC", seq_along(share)),
        ylim = c(0, 1), ylab = "Fraction of weighted variation",
        col = "grey60", border = NA)
```

The first two components account for
`r sprintf("%.1f%%", 100 * sum(share[1:2]))` of the weighted variation in this
example. A scree plot describes the training fit; it does not establish how
well components will generalize. For prediction, compare held-out
reconstruction error using preprocessing and data-derived metrics estimated
on training rows only.

## Where next

- [GPCA Metrics](gpca-metrics.html): construct row and column metrics and choose
  between smoothing and noise whitening.
- [Modelling Structured Noise](structured-noise.html): explore those choices
  with known signal and noise in simulations.
- [GPCA at Scale](gpca-scale.html): choose a backend, budget memory, and project
  held-out observations.
- [Sparse and Functional PCA](sfpca.html): fit sparse, smooth factors.
- [Generalized PLS-SVD](gplssvd-reference.html): relate two data blocks, with an
  explicit whitening reference for contributors.

The remaining sections explain the mathematics and notation behind the fit.

## The decomposition

### Two metrics, one weighted objective

Let $X$ be the $n \times p$ data matrix. Generalized PCA asks you to supply
two symmetric positive semi-definite matrices:

- $M$ ($n \times n$), the **row metric**, which puts a geometry on the space of observations;
- $A$ ($p \times p$), the **column metric**, which puts a geometry on the space of variables.

For positive definite metrics, they define an inner product on $n \times p$ matrices,

$$\langle Y, Z \rangle_{M,A} \;=\; \operatorname{tr}\!\left(Y^{\top} M Z A\right),
\qquad
\|Y\|_{M,A}^{2} \;=\; \operatorname{tr}\!\left(Y^{\top} M Y A\right).$$

When $M = I_n$ and $A = I_p$ this is the ordinary Frobenius inner product,
and the decomposition reduces to the familiar SVD and PCA. Singular PSD metrics instead define
a seminorm: directions in their null spaces have zero weight. Here $X$ denotes
the preprocessed data, after centering if requested.

### The problem it solves

The **generalized least-squares matrix decomposition** (GMD) of Allen,
Grosenick & Taylor (2014) is the best rank-$K$ approximation of $X$ *in that
norm*:

$$\min_{U, D, V} \;\bigl\|X - U D V^{\top}\bigr\|_{M,A}^{2}
\qquad \text{subject to} \qquad
U^{\top} M U = I_K, \quad V^{\top} A V = I_K,$$

with $D = \operatorname{diag}(d_1 \ge d_2 \ge \cdots \ge d_K \ge 0)$.

Only two things changed relative to the SVD, and they are the same change
twice: the discrepancy is measured in the $(M,A)$ norm rather than the
Frobenius norm, and the factors are orthonormal *in the metrics* rather than
Euclidean-orthonormal.

### The solution: whiten, decompose, unwhiten

The problem has a closed form. Take symmetric square roots $M = M^{1/2}M^{1/2}$
and $A = A^{1/2}A^{1/2}$, whiten the data,

$$\widetilde{X} \;=\; M^{1/2} X A^{1/2},$$

and compute its *ordinary* SVD, $\widetilde{X} = \widetilde{U} D \widetilde{V}^{\top}$.
Then

$$U = M^{-1/2}\widetilde{U}, \qquad V = A^{-1/2}\widetilde{V}, \qquad D \text{ unchanged}$$

solves the problem above (with pseudo-inverses when a metric is singular).
GMD is therefore an SVD in the coordinates selected by the metrics.

For positive definite metrics, eliminating the whitening gives the eigenproblem

$$X^{\top} M X A \, v_k \;=\; d_k^{2}\, v_k, \qquad v_k^{\top} A v_k = 1,$$

and the $k$-th generalized principal component (the score vector) is

$$z_k \;=\; X A v_k \;=\; d_k\, u_k .$$

For singular metrics, the pseudo-inverse factors solve the problem on the
metrics' supported subspaces; the identities above need the corresponding
projections when interpreted outside those subspaces.

In the package these are `components(fit)` $= A\,$`ov` for the loadings and
`scores(fit)` $=$ `ou %*% diag(sdev)` for the scores, with `ou` and `ov`
holding the metric-orthonormal factors $U$ and $V$.

### What the metrics mean

There are two readings of $M$ and $A$, and both are worth carrying.

**As geometry.** The metrics define what "distance" means, and they act
crosswise. Observations are rows of $X$, living in $\mathbb{R}^p$, so the
*column* metric sets the distance between two observations,
$\|x_i - x_j\|_A^2 = (x_i - x_j)^{\top} A\,(x_i - x_j)$. Variables are
columns, living in $\mathbb{R}^n$, so the *row* metric sets the distance
between two variables — and, when $M$ is diagonal, simply weights the
observations. Choose $A$ to say which variables should be treated as similar;
choose $M$ to say which observations count more.

**As statistics.** Suppose the noise is separable, meaning the errors follow a
matrix-normal law $E \sim \mathcal{MN}(0, \Sigma_{\text{row}}, \Sigma_{\text{col}})$,
equivalently $\operatorname{vec}(E)$ has covariance
$\Sigma_{\text{col}} \otimes \Sigma_{\text{row}}$. Then the log-likelihood of a
rank-$K$ mean is, up to constants, exactly $-\tfrac12\|X - UDV^{\top}\|^2_{M,A}$
with

$$M = \Sigma_{\text{row}}^{-1}, \qquad A = \Sigma_{\text{col}}^{-1}.$$

So the metrics are not an arbitrary regularizer: setting them to the inverse
noise covariances makes GMD the maximum-likelihood low-rank fit. This is why
the package ships `gpca_mle()`, which learns $(M, A)$ by penalized maximum
likelihood, and `mnpca_mrl()`, which does the same with sparse precision
matrices.

## Choosing structured metrics

Use [GPCA Metrics: Building M and A](gpca-metrics.html) for runnable recipes.
The direction of the weighting matters:

- To favour smooth column loadings, start with a PSD kernel or a smoother
  such as $A = (I + \alpha L)^{-1}$, where $L$ is a graph Laplacian.
  A raw adjacency or distance matrix need not be PSD and cannot be used
  without a valid metric construction.
- To account for correlated noise, use its **inverse covariance**. For a
  time-by-voxel table with separable Gaussian noise, choose
  $M = \Sigma_{\mathrm{time}}^{-1}$ and
  $A = \Sigma_{\mathrm{space}}^{-1}$. The row metric acts on time and the
  column metric on space. A covariance model for related samples likewise
  enters through its inverse when the goal is noise whitening.
- `sfpca()` encourages smooth, sparse factors through a different constraint.
  Its spatial penalty is built from `spat_cds`; increasing `alpha_v`
  strengthens that penalty. See [Sparse and Functional PCA](sfpca.html)
  before transferring tuning choices between the two interfaces.

A Laplacian as `A` emphasizes contrasts across graph edges; its inverse
(with a ridge to make it invertible) emphasizes smooth directions. Select
between these according to whether smooth variation is signal or nuisance.

## Choosing a backend

The default `method = "eigen"` uses a Gram eigenproblem on the smaller side
of the data. `"spectra"` usually applies eigencore's partial SVD to a
whitened operator, with a dense fallback and a small-side Gram route for a
singular large-side metric. `"randomized"` approximates a leading subspace;
`"deflation"` extracts components one at a time and can preserve sparse data.

Neither `"spectra"` nor `"randomized"` preserves sparse `X`: both make a
dense data copy. Metric factorizations and fallback storage also matter.
[GPCA at Scale](gpca-scale.html) describes those costs, the `maxeig` guard,
and the covariance-only interface.

## What are `ou`, `ov`, `u`, and `v`?

Under the hood, `genpca()` stores the raw factors of the decomposition
alongside the verbs above. `ou` and `ov` are the metric-orthonormal factors
of the paper (Allen, Grosenick & Taylor 2014): `ou` is orthonormal in the
row metric (`t(ou) %*% M %*% ou` is the identity) and `ov` is orthonormal in
the column metric (`t(ov) %*% A %*% ov` is the identity). `u` and `v` are
the same factors after applying the metric (`u = M %*% ou`, `v = A %*% ov`),
which is the form used for `components()`. The table below ties these to
the verbs:

| Quantity | Definition | Relation to verbs |
|:--|:--|:--|
| `ou` | Row-metric-orthonormal left factor | `t(ou) %*% M %*% ou = I` |
| `ov` | Column-metric-orthonormal right factor | `t(ov) %*% A %*% ov = I` |
| `v` (`components(fit)`) | `A %*% ov` | Loadings in the original variable space |
| `s` (`scores(fit)`) | `ou %*% diag(sdev)` | Stored training scores |

Here the decomposition applies to the preprocessed matrix `Xp`, not raw `X`.
For positive definite metrics, the score formula is `Xp %*% A %*% ov`.
`project()` applies the stored preprocessing before multiplying by the
projection weights. With singular metrics, null-space representatives can
make projected coordinates differ from the stored factor scores outside the
weighted subspace.

## Relation to the French school

Readers coming from *analyse des données* will recognize all of this. The
French tradition — Benzécri's correspondence analysis, formalized by
Escoufier as the **duality diagram** — describes an analysis by a *triplet*
$(X, Q, D)$: a data table, a metric on the variable space, and a set of row
masses. The analysis is then the eigendecomposition of the operator
$X^{\top} D X Q$. That is, term for term, the eigenproblem in [The decomposition](#the-decomposition). Correspondence analysis, multiple correspondence analysis, principal
coordinates, discriminant analysis, canonical correlation, and PLS are all
recovered by choosing the triplet appropriately; the `ade4` package is built
directly on this formalism, and Abdi's papers on the generalized SVD — cited
in this package's `DESCRIPTION` — are the same construction in English.

Watch the notation: **$Q$ means opposite things in the two
literatures.**

| Role | `genpca` | Allen et al. (2014) | Duality diagram |
|:--|:--|:--|:--|
| Row metric / observation weights | `M` | $Q$ | $D$ |
| Column metric / variable geometry | `A` | $R$ | $Q$ |

What Allen, Grosenick & Taylor contribute on top of the classical theory is
less the geometry than the framing and the machinery: the decomposition is
posed as a *best rank-$K$ approximation* with an optimality theorem rather
than only an eigenanalysis; the metrics are allowed to be large, dense, and
structured (spatial kernels, temporal covariances) instead of the diagonal
weights typical of the classical treatments; and the formulation extends
cleanly to regularized and sparse variants, which is what `sfpca()`
implements.

## References

Allen, G. I., Grosenick, L., & Taylor, J. (2014). A generalized least-square
matrix decomposition. *Journal of the American Statistical Association*,
109(505), 145–159.

Escoufier, Y. (1987). The duality diagram: a means for better practical
applications. In *Development in Numerical Ecology*. Springer.

Holmes, S. (2008). Multivariate data analysis: the French way. In
*Probability and Statistics: Essays in Honor of David A. Freedman*. IMS.

Dray, S., & Dufour, A.-B. (2007). The ade4 package: implementing the duality
diagram for ecologists. *Journal of Statistical Software*, 22(4).

Silverman, B. W. (1996). Smoothed functional principal components analysis by
choice of norm. *The Annals of Statistics*, 24(1), 1–24.
