---
title: 'GPCA Metrics: Building M and A'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{GPCA Metrics: Building M and A}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  message    = FALSE,
  warning    = TRUE,
  fig.width  = 6,
  fig.height = 4,
  out.width  = "85%"
)
library(genpca)
library(Matrix)

flip <- function(M) as.matrix(M)[, ncol(M):1]
heat <- function(M, main = "") {
  image(flip(M), axes = FALSE, main = main,
        col = grey.colors(20, start = 0.95, end = 0.2))
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```




This vignette collects practical recipes for row and column metrics, plus notes on SPD remedies and the experimental `gpca_mle()` learner.

## Why metrics

Metrics encode weighting and correlation. The row metric `M` changes how
observations are compared; the column metric `A` changes how variables are
compared. Setting either to something other than the identity is how you tell
the decomposition what you already know about the data: that the samples are a
time series, that the variables sit on a spatial grid or fall into groups, that
some measurements are noisier than others. Ordinary PCA has no way to accept
that information and treats every row and column alike. GPCA writes it into the
objective being optimised.

Structure alone is not enough, though: you also have to get the *direction*
right, and supplying a matrix versus its inverse produces opposite results.
The next section is about that, and it is worth reading before the recipes.

Both metrics must also be symmetric and positive semi-definite. That rarely
gets in the way, but it does constrain what you can pass; see
[SPD requirements and remedies](#spd-requirements-and-remedies) below for what
the requirement means and what `genpca()` does when a metric falls short of it.

## Heteroscedastic diagonals

The simplest non-trivial metric is a diagonal that down-weights noisy rows or columns:

```{r hetero-demo}
set.seed(42)
n <- 60; p <- 20
X <- matrix(rnorm(n * p), n, p)

col_noise_sd <- runif(p, 0.5, 2)
A <- Diagonal(x = 1 / col_noise_sd^2)
row_noise_sd <- runif(n, 0.7, 1.3)
M <- Diagonal(x = 1 / row_noise_sd^2)

fit <- genpca(X, M = M, A = A, ncomp = 3,
              preproc = multivarious::center())
fit$sdev
```

```{r hetero-plot, echo = FALSE, fig.cap = "Inverse-variance weights on columns (top) and rows (bottom). Noisier dimensions get smaller weights.", fig.height = 3.5}
op <- par(mfrow = c(2, 1), mar = c(2.5, 4, 2, 1))
barplot(diag(A), border = NA, col = "steelblue",
        main = "Column weights diag(A)", names.arg = NA)
barplot(diag(M), border = NA, col = "tomato",
        main = "Row weights diag(M)", names.arg = NA)
par(op)
```

### Inverse column variance *is* scaled PCA

Inverse-variance weighting on the columns is not a new idea in disguise: it is
exactly the standardisation that `prcomp(scale. = TRUE)` performs. GPCA whitens
with the square root $A^{1/2}$, so setting $A = \operatorname{diag}(1/s_j^2)$
makes $X A^{1/2} = X \operatorname{diag}(1/s_j)$ — the column-scaled matrix
that correlation-matrix PCA decomposes.

```{r scaled-pca-equiv}
set.seed(42)
Xv <- matrix(rnorm(60 * 20), 60, 20) %*% diag(runif(20, 0.5, 3))
sds <- apply(Xv, 2, sd)

g  <- genpca(Xv, A = Diagonal(x = 1 / sds^2), ncomp = 5,
             preproc = multivarious::center())
pr <- prcomp(Xv, scale. = TRUE)

# scores agree component by component
sapply(1:3, function(k) cor(multivarious::scores(g)[, k], pr$x[, k]))
```

The scores are identical (up to sign). The reported `sdev` values differ by a
single constant, because `prcomp()` divides its singular values by
$\sqrt{n-1}$ and `genpca()` reports them unnormalised:

```{r scaled-pca-sdev}
rbind(genpca = g$sdev[1:5],
      prcomp = pr$sdev[1:5],
      ratio  = g$sdev[1:5] / pr$sdev[1:5])
sqrt(nrow(Xv) - 1)
```

The constant ratio is the whole difference. This is worth internalising as the
baseline: **a diagonal `A` generalises column scaling**, and everything else in
this vignette — kernels, Laplacians, AR(1) structure — generalises it further
by letting the metric go off-diagonal.

### Weighting both margins at once

The example above sets `M` and `A` simultaneously, which is a reasonable thing
to want when both observations and variables are heteroscedastic. Two
consequences are easy to miss.

**The two weightings interact.** GPCA works with $M^{1/2} X A^{1/2}$, so the
row weights change each column's effective variance and the column weights
change each row's. Estimating row and column standard deviations from the raw
data and applying both at once therefore standardises *neither* margin:

```{r both-margins}
Xc <- scale(Xv, center = TRUE, scale = FALSE)
W  <- diag(1 / apply(Xc, 1, sd)) %*% Xc %*% diag(1 / apply(Xc, 2, sd))

range(apply(W, 1, sd))   # row SDs, would be constant if standardised
range(apply(W, 2, sd))   # column SDs
```

Each rescaling changes the other margin's standard deviations, so one pass
does not produce unit variance on both. Alternating marginal scaling is a
separate procedure from `gpca_mle()`: that learner alternates a low-rank fit
with estimates of the full residual row and column covariances, including a
ridge penalty. It does not simply standardize the two margins.

**Only the product of the two scales is identified.** Replacing $(M, A)$ with
$(cM, A/c)$ leaves $M^{1/2} X A^{1/2}$ untouched, so the fit cannot distinguish
them:

```{r scale-indeterminacy}
M0 <- Diagonal(x = 1 / apply(Xv, 1, sd)^2)
A0 <- Diagonal(x = 1 / sds^2)

f1 <- genpca(Xv, M = M0,     A = A0,     ncomp = 4, preproc = multivarious::center())
f2 <- genpca(Xv, M = 7 * M0, A = A0 / 7, ncomp = 4, preproc = multivarious::center())

max(abs(f1$sdev - f2$sdev))
```

The singular values and the component subspace are identical; only the scores
pick up a constant factor, since their normalisation is tied to the scale of
`M`. The practical upshot is that there is no point tuning the overall
magnitude of `M` against that of `A` — it is the *relative* weighting within
each metric that changes the answer. This indeterminacy is also why
`gpca_mle()` has a `scale_fix` argument: when both metrics are learned, the
split is pinned down by the ridge penalty in the objective (the default,
`scale_fix = "none"`), and `"trace"`/`"det"` are optional post-hoc
reparameterizations whose effect on the penalized objective is reported in
`loglik_rescale_delta`.

## Which way does a metric point?

Before the recipes, the single most important thing to get right — and the
easiest to get backwards. **A metric amplifies its own dominant
eigendirections.**

Seeing why means being precise about what comes back from a fit. GPCA
factorises $X \approx U D V^{\top}$, where $V$ is orthonormal *in the column
metric* rather than in the ordinary sense: $V^{\top} A V = I$. The loadings
returned by `components(fit)` are not $V$ but $AV$:

```{r loadings-are-AV}
set.seed(1)
Xd <- matrix(rnorm(400), 40, 10)
Ad <- crossprod(matrix(rnorm(100), 10, 10)) / 10 + diag(10)
fd <- genpca(Xd, A = Ad, ncomp = 3, preproc = multivarious::center())

max(abs(multivarious::components(fd) - as.matrix(Ad %*% fd$ov)))
```

Multiplication by $A$ changes how the fitted factor is expressed. It stretches every direction in
proportion to the eigenvalue $A$ assigns it, so the patterns $A$ scores highly
are the patterns that dominate the loadings you read off. (The bare factor $V$
is kept in the `ov` slot if you ever need it, but `components()` is what you
should normally interpret.)

For a spatial or temporal structure there are two natural matrices, and they
point in *opposite* directions:

| You supply | $v^{\top}Av$ measures | Large eigenvalues on | Components come out |
|:--|:--|:--|:--|
| A **smoother**: kernel $K$, adjacency $I + \alpha W$, $(I+\alpha L)^{-1}$, heat kernel $e^{-tL}$ | agreement between neighbours | smooth patterns | **smoother** |
| A **precision**: Laplacian $L$, $I + \alpha L$, $K^{-1}$, inverse AR(1) | disagreement across edges (Dirichlet energy $\sum_{i\sim j}(v_i - v_j)^2$) | rough patterns | **rougher** |

Both are legitimate, because they encode different beliefs about *where the
noise lives*. The bridge is $A = \Sigma_{\text{col}}^{-1}$, and the step that
is easy to skip is the inversion: the metric and the noise covariance share
eigenvectors but carry **reciprocal** weights, so a direction the metric scores
highly is a direction the noise model calls quiet.

That reciprocal is worth seeing rather than taking on faith. On a cycle graph
the Laplacian and the adjacency share Fourier eigenvectors exactly, so
"roughness" is unambiguously frequency:

```{r reciprocal-weights}
p <- 32
Wc <- matrix(0, p, p)
for (i in 1:p) { Wc[i, i %% p + 1] <- 1; Wc[i %% p + 1, i] <- 1 }
Lc <- diag(rowSums(Wc)) - Wc

ec <- eigen(Lc, symmetric = TRUE)
o  <- order(ec$values)
Vc <- ec$vectors[, o]          # smoothest first
lc <- ec$values[o]             # Dirichlet energy = roughness

Ac  <- diag(p) + 0.45 * Wc     # a smoother (PSD)
Sig <- solve(Ac)               # the noise covariance it implies

modes <- c(1, 16, 32)          # smoothest, middling, roughest
data.frame(
  roughness   = round(lc[modes], 3),
  metric_wt   = round(sapply(modes, function(k) t(Vc[, k]) %*% Ac  %*% Vc[, k]), 3),
  noise_var   = round(sapply(modes, function(k) t(Vc[, k]) %*% Sig %*% Vc[, k]), 3)
)
```

The metric weight *falls* with roughness while the implied noise variance
*rises*, each the reciprocal of the other. Note this is a statement about
$\Sigma_{\text{col}} = A^{-1}$, not about the adjacency matrix you supplied:
$A$ itself has its large eigenvalues on the **smooth** directions. Inverting
is what moves the variance to the rough end.

- **Smoother as metric** $\Rightarrow$ $A$ is large on smooth directions
  $\Rightarrow$ $\Sigma_{\text{col}} = A^{-1}$ is large on the rough ones
  $\Rightarrow$ "the noise is high-frequency speckle; the signal is smooth."
  This is denoising, and it is what you want when you are after spatially
  coherent maps.
- **Precision as metric** $\Rightarrow$ $A$ is large on rough directions
  $\Rightarrow$ $\Sigma_{\text{col}}$ is large on the smooth ones $\Rightarrow$
  "a smooth field — drift, a scanner gradient, a global trend — is the
  nuisance." Whitening it away lets fine-scale structure surface. This is
  ordinary generalized least squares against correlated noise.

One caveat on how literally to read this. $A = \Sigma_{\text{col}}^{-1}$ is an
interpretive frame, not a constraint `genpca()` enforces — the algorithm only
ever whitens with $A^{1/2}$. The covariance reading is how you should *choose*
a metric, and it is what makes "smoother $\Rightarrow$ denoising" more than a
slogan.

So the question is never "adjacency or Laplacian?" in the abstract. It is:
**is the smooth thing my signal, or my nuisance?**

```{r orientation-demo, echo = FALSE, fig.cap = "The same graph, two metrics. Left: a smoother concentrates PC1 on the smooth blob. Right: the Laplacian whitens the smooth field away, so PC1 locks onto the fine-scale checkerboard that was buried underneath it.", fig.width = 7, fig.height = 3.5}
set.seed(7)
gg <- 12; pp <- gg * gg; nn <- 80
gidx <- expand.grid(r = 1:gg, c = 1:gg)
Wg <- matrix(0, pp, pp)
for (i in 1:pp) for (j in 1:pp)
  if (i < j && abs(gidx$r[i] - gidx$r[j]) + abs(gidx$c[i] - gidx$c[j]) == 1) {
    Wg[i, j] <- 1; Wg[j, i] <- 1
  }
Lg <- diag(rowSums(Wg)) - Wg
blob <- exp(-((gidx$r - 4)^2 + (gidx$c - 4)^2) / 6); blob <- blob / sqrt(sum(blob^2))
chk  <- as.numeric(((gidx$r + gidx$c) %% 2) * 2 - 1); chk <- chk / sqrt(sum(chk^2))
Xg <- matrix(rnorm(nn), nn, 1) %*% t(chk) * 3 +
      matrix(rnorm(nn), nn, 1) %*% t(blob) * 15 +
      matrix(rnorm(nn * pp), nn, pp) * 0.5
f_sm  <- genpca(Xg, A = solve(diag(pp) + 2 * Lg), ncomp = 1, preproc = multivarious::center())
f_lap <- genpca(Xg, A = diag(pp) + 50 * Lg,      ncomp = 1, preproc = multivarious::center())
op <- par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
for (ff in list(list(f_sm, "Smoother metric: finds the blob"),
                list(f_lap, "Laplacian metric: finds the checkerboard"))) {
  v <- multivarious::components(ff[[1]])[, 1]
  image(matrix(v, gg, gg), axes = FALSE, main = ff[[2]], cex.main = 0.95,
        col = grey.colors(24, start = 0.95, end = 0.15))
}
par(op)
```

Two practical notes on the graph matrices themselves. A raw adjacency $W$ is
**indefinite** (its eigenvalues sum to zero), so it is not a valid metric on
its own — shift it, as in $I + \alpha W$ with $\alpha$ small enough to keep
it PSD. A raw Laplacian $L$ is PSD but **singular**: $L\mathbf{1} = 0$, so
the spatially constant pattern has zero length under it. `genpca()` accepts
that (the whitening uses a pseudo-inverse), but adding a small ridge,
$L + \varepsilon I$, is usually what you want.

The example compares a smoother with coefficient 2 against a precision with
coefficient 50. It demonstrates these two choices, not a universal switching
threshold. Sweep the strength on your own data and inspect the loadings.

## Recipes

Each recipe below is labelled with the direction it produces. All three are
written as **precision** matrices. For the AR(1) and RBF recipes, use the
covariance inside `solve()` to obtain the smoother orientation. The Laplacian
recipe constructs a precision directly; invert its regularized matrix to
obtain a smoother.

### AR(1) row metric (whitens temporal autocorrelation)

Standard GLS treatment of serially correlated observations, as in fMRI
prewhitening: it removes temporal autocorrelation rather than imposing
temporal smoothness.

```{r ar1}
rho     <- 0.7
n_t     <- 60
idx     <- 0:(n_t - 1)
Sigma_r <- outer(idx, idx, function(i, j) rho^abs(i - j))
M_ar1   <- solve(Sigma_r + 1e-3 * diag(n_t))
```

### Spatial RBF kernel (whitens smooth spatial noise)

```{r rbf}
coords <- as.matrix(expand.grid(x = 1:8, y = 1:8))
d2     <- as.matrix(dist(coords))^2
ell    <- 2
K      <- exp(-d2 / (2 * ell^2))
A_rbf  <- solve(K + 1e-3 * diag(nrow(K)))   # precision: emphasises fine scale
# A_smooth <- K + 1e-3 * diag(nrow(K))      # kernel itself: smooth loadings
```

### Graph Laplacian (emphasises contrast across edges)

```{r lap}
W <- bandSparse(30, k = c(-1, 0, 1),
                diagonals = list(rep(0.2, 29),
                                 rep(1, 30),
                                 rep(0.2, 29)))
D     <- Diagonal(x = rowSums(W))
A_lap <- (D - W) + 1e-2 * Diagonal(nrow(W))
# A_smooth <- solve(A_lap)                  # smoother: spatially coherent loadings
```

```{r recipe-plots, echo = FALSE, fig.cap = "Three structured metrics, all shown in the precision (noise-whitening) orientation. Off-diagonal banding is what couples nearby rows or variables -- it tells GPCA 'treat these dimensions as related, not independent'.", fig.width = 8, fig.height = 3}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
heat(M_ar1, "Inverse AR(1)")
heat(A_rbf, "Inverse RBF kernel")
heat(A_lap, "Regularised Laplacian")
par(op)
```

### A note on `sfpca()`

`sfpca()` takes the *opposite* input for the same intent. There the structure
enters as a constraint, $v^{\top}(I + \alpha\Omega)v \le 1$, which charges
rough $v$ against a fixed budget — so the spatial roughness operator is built from `spat_cds`, and `alpha_v`
controls its strength. Larger `alpha_v` means **smoother**; it is a scalar,
not an argument for supplying a matrix. Metric form and constraint form are inverse to one
another: the same Laplacian smooths in `sfpca()` and roughens in `genpca()`.

## Learning metrics with `gpca_mle()`

`gpca_mle()` is an experimental learner that alternates a low-rank fit with
regularized matrix-normal covariance estimates. Use it to explore estimated
metrics, checking their spectra and sensitivity to the ridge parameter
`lambda`. An i.i.d. input does not guarantee an identity-like fitted metric:
a single small data matrix provides limited information about unrestricted
row and column covariances, especially after fitting a low-rank mean.

```{r mle-fit}
set.seed(1)
n_m <- 40; p_m <- 10
X_mle <- matrix(rnorm(n_m * p_m), n_m, p_m)
fit_mle <- gpca_mle(X_mle, ncomp = 2, max_iter = 6,
                    lambda = 1e-3, scale_fix = "none",
                    method = "eigen", verbose = FALSE)

metric_spectrum <- function(W) {
  ev <- eigen(as.matrix(W), symmetric = TRUE, only.values = TRUE)$values
  c(min = min(ev), max = max(ev), condition = max(ev) / min(ev))
}
signif(rbind(M = metric_spectrum(fit_mle$M),
             A = metric_spectrum(fit_mle$A)), 3)
```

The row metric has widely separated eigenvalues. A heatmap can look nearly
diagonal while hiding this distinction, so plot the spectrum after removing
overall scale. An identity-like metric would have every normalized eigenvalue
near one.

```{r mle-plot, echo = FALSE, fig.cap = "Metric eigenvalues divided by their mean, on a log scale. The dashed line marks an identity-like spectrum; the learned row metric departs strongly from it.", fig.width = 7, fig.height = 3.5}
spectra <- lapply(list(M = fit_mle$M, A = fit_mle$A), function(W) {
  ev <- eigen(as.matrix(W), symmetric = TRUE, only.values = TRUE)$values
  ev / mean(ev)
})
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (nm in names(spectra)) {
  plot(spectra[[nm]], type = "b", pch = 19, log = "y",
       ylim = range(unlist(spectra)), xlab = "Eigenvalue index",
       ylab = "Eigenvalue / mean", main = paste("Learned", nm))
  abline(h = 1, lty = 2, col = "steelblue")
}
par(op)
```

```{r mle-spectrum-check, include = FALSE}
stopifnot(all(is.finite(unlist(spectra))), all(unlist(spectra) > 0))
```

Inspect optimization progress as well:

```{r mle-progress}
data.frame(iteration = seq_along(fit_mle$loglik_path),
           penalized_loglik = fit_mle$loglik_path)
```

Six iterations are an illustration, not evidence of convergence or covariance
recovery. Keep `lambda` positive and compare results across its plausible
values. The default `scale_fix = "none"` retains the scale selected by the
penalized fit; optional `"trace"` or `"det"` rescaling changes the penalized
objective, reported in `loglik_rescale_delta`. The separate `loglik_refit_delta`
records the remaining difference between the last iteration's objective and
the returned objective, including the final refit and numerical reevaluation.
With `scale_fix = "none"`, the rescale delta is exactly zero even when the
refit delta is nonzero.

## SPD requirements and remedies

### Why the requirement exists

GPCA measures squared lengths using $u^\top M u$ and $v^\top A v$,
and the solvers whiten the data
with the square roots $M^{1/2}$ and $A^{1/2}$. Both steps need the metrics to
be symmetric positive semi-definite (PSD). If a metric has a negative
eigenvalue, vectors in that direction have negative squared length, the square
root is not real, and "maximise variance" no longer picks out anything
meaningful.

Note that *semi*-definite is enough. **Singular metrics are perfectly legal.**
A graph Laplacian is rank-deficient by construction — it has a zero eigenvalue
on the constant vector — and `genpca()` takes it without complaint. A zero
eigenvalue simply means that direction is given no weight. Only *negative*
eigenvalues are a problem.

### What happens when a metric falls short

Distinguish singularity from invalidity: a sample covariance from fewer
samples than variables can be singular and still PSD. Rounding can produce
small negative eigenvalues or slight asymmetry; an incorrectly constructed
metric can produce larger violations. The checks are relative to
the scale of the matrix, with a tolerance for floating-point noise: eigenvalues down to $-\sqrt{\epsilon}\,\max|A_{ii}|$ (about
$-1.5\times10^{-8}$ times the largest diagonal entry) count as non-negative, and
an asymmetry $\|A - A^\top\|_F / \|A\|_F$ below $10^{-10}$ is averaged away.
Tiny asymmetry may therefore be averaged away; metric factorization also
uses numerical tolerances to identify null directions. An explicit `"clip"`
request removes negative eigenvalues even if they pass the tolerant PSD check.

A genuinely asymmetric matrix is an error under
every setting: there is no way to know which triangle you meant. And a metric
that fails the PSD check is an error by default (`constraints_remedy =
"error"`), because a fit that silently ran on a different metric than the one
you supplied is worse than no fit. If you do want a repair, ask for it, and
you will be told what was done:

| Value | What it does | What it costs |
|---|---|---|
| `"error"` (default) | Refuses the input. | Nothing — this is the right setting when the metric comes from a pipeline that ought to be producing a valid one. |
| `"ridge"` | Adds a diagonal shift (from the Gershgorin bound, with a `Matrix::nearPD()` fallback for small dense matrices) sufficient to make the matrix positive definite. Preserves sparsity. | The shift pulls the metric toward a multiple of the identity, diluting the structure you supplied. A *large* shift means the input was badly indefinite — diagnose it rather than absorb it. |
| `"clip"` | Eigendecomposes and sets the negative eigenvalues to zero, leaving the rest of the spectrum exactly as it was. | Densifies the matrix, so it refuses sparse input larger than 2000×2000. Use `"ridge"` at that size. |
| `"identity"` | Replaces the offending metric with the identity. | Discards the offending metric; the other metric still applies. |

Every repair that actually changes the metric emits a warning of class
`genpca_metric_repaired` whose `report` field records the minimum eigenvalue
before and after, the shift applied, the rank and the condition number. The
same report is available directly from `repair_metric()`, which is the better
way to work: repair once, look at the report, and pass the repaired matrix to
every subsequent fit.

```{r spd-check, eval = FALSE}
# Is the metric usable as-is, and what would a repair do to it?
A_ok <- repair_metric(A, method = "ridge")
attr(A_ok, "repair_report")

# Catch the repair warning programmatically inside a fit
fit <- withCallingHandlers(
  genpca(X, A = A, M = M, ncomp = 3, constraints_remedy = "ridge"),
  genpca_metric_repaired = function(w) { print(w$report); invokeRestart("muffleWarning") }
)
```

Prefer metrics that are PSD by construction, such as a PSD kernel, a graph
Laplacian, or a nonnegative diagonal. Dividing a metric by its mean diagonal
changes its overall magnitude but leaves its condition number unchanged.
Inspect the repair report when a repair is requested, and inspect learned
metric spectra even when no warning is emitted.

## Where next

See [Modelling Structured Noise](structured-noise.html) for how to choose the transfer function when several kinds of structure are present at once, and [GPCA at Scale](gpca-scale.html) for backend choices, sparse workflows, and covariance-only GPCA.
